Gaia's constraints on the very massive $\sim$$150\,M_\odot$ runaway star candidate VFTS682

### 2018-07-22 VFTS682_APJL.tex
added ignore comments folder
included Steve Shore's comments

% Aiming to shine light on this debate,
In this context, 
we investigate the kinematics of
VFTS682 as obtained by \emph{Gaia} and multi-epoch \emph{Hubble Space
  Telescope} photometry. We derive a projected velocity relative of to
the cluster of $\sim$$30\pm20\kms$.  The direction and magnitude of
the proper motion are consistent with what is expected if VFTS682 was ejected as a slow runaway from the central cluster, given its distance and the inferred age.  However, the error bars are substantial and the proper motion measurements alone cannot rule out the counter hypothesis with high confidence.

@@

The WNh5 star VFTS682, located at right ascension (RA)
05$^\mathrm{h}$38$^\mathrm{m}$55.510$^\mathrm{s}$  and declination
(DEC) \mbox{-69$^\mathrm{o}$04'26.72''} J2000, was observed as part of the multi-epoch, spectroscopic VFTS campaign covering $\lambda$4000--7000 \citep[][]{evans:11}. 
\citet{bestenlehner:11}  analyzed the spectra to infer the stellar
parameters and measure an extinction of $A_V=4.45\pm0.12$, implying a
luminosity of $\log_{10}(L/L_\odot) =  6.5\pm0.2$, making this one of
the brightest stars in the region. The star is unlikely to have a
close companion, because of the absence of clear radial velocity
variations \citet{bestenlehner:11}. Bayesian fits of the stellar
parameters against evolutionary tracks \citep{brott:11, kohler:15}
using the BONNSAI code \citep{schneider:17, schneider:18} provide
estimates for the age, present mass and initial mass, % indicating that
% this star among the most massive stars known,
% see
\Tabref{tab:star_param}. % for an overview of the parameters.

@@

\cite{townsley:06} and \textcolor{blue}{Townsley et al., in prep.} did not find
an X-ray point source at the position of VFTS682 in their
\emph{Chandra} survey of the region, which suggests the absence of
colliding winds that would be expected in the presence of companions
even for extreme mass ratios, given the extremely large mass of
VFTS682. This star is also relatively isolated in (near) infrared. There nearest bright (near-)infrared sources detected by \emph{Spitzer} \citep{meixner:06} and resolved in the \emph{VISTA} Magellanic Clouds Survey \citep{cioni:11} are located at a distance of about 10 arcsecond, i.e. about 2.4 pc. \cite{walborn:13} speculate that these nearby young stars may represent a case of starformation triggered by the the wind of VFTS682, but we chance alignment cannot be excluded.

@@

We adopt as distance to the LMC the value of $50$\,kpc. The error on the distance determination is small ($\lesssim2\%$, \citealt{pietrzynski:13}) and any possible offset between  R136 or VFTS682  and the distance we adopted for the LMC is probably much smaller (likely $\ll 1\%$, e.g., \citealt{Luks+1992}). These uncertainties are negligible compared to the errors in the proper motion discussed below.

@@

\vspace*{10pt}% to move new sec. to newpage

@@

\begin{figure*}[tbp]
  \centering
  \includegraphics[width=0.35\textwidth]{figures/dist_mu_region.pdf}
  \includegraphics[width=0.35\textwidth]{figures/angle}
  \caption{\emph{Left panel}: distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136. VFTS682 is not an outlier, but
    its relative proper motion matches the expected value if it were indeed
    ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. The top axis shows the conversion to physical units
    assuming a distance of 50\,kpc. \emph{Right panel}:  distribution of
    angles between the relative proper motion direction and the radial
    direction from the center of the cluster to the star. The error bars from \emph{Gaia} for VFTS682 are large, but
    the best value is in agreement with the hypothesis of dynamical
    ejection. The peak at $\theta\simeq\pi/2$ is due to stars
    belonging to NGC2060. In both
    panels, the dark blue histograms contain 317 
    stars with error smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
    lighter blue histograms contain 36 stars with errors smaller than $0.05\,\mathrm{mas \
      yr^{-1}}$. }
  \label{fig:dist}
\end{figure*}

@@

Assuming VFTS682 indeed originates from R136, we can calculate its kinematic
age simply as:
\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{\delta\mu_{Gaia}} \simeq
  \frac{119.4\,\mathrm{arcsec}}{0.13\masyr} \simeq 0.9\pm\,0.6\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel = 119.4\,\mathrm{arcsec}$ is the angular distance from VFTS682 to
the core of the cluster \citep[corresponding to $\sim$29\,pc at a
distance of 
50\,kpc,][]{bestenlehner:11}.
% As in the rest of this study, we neglect for
% simplicity the error on the distance estimates, because it is negligible compared to other uncertainties.
The kinematic age $\tau_\mathrm{kin}$ is compatible with a very early
ejection from the cluster, given the apparent age of the star.
% of $1.0\pm 0.2$\,Myr 
% \citep{schneider:18}.

@@

% \subsection{Is it a runaway star?}
% \label{sec:runaway}
% We first address the question whether VFTS682 is a typical star
% from the kinematic point of view (which is what would be expected if it formed in relative isolation), or whether has a significant peculiar velocity compared to its surrounding population (which is what we would expect if the star has a runaway origin).

% After subtracting the mean proper motion of the field, we obtain the relative two-dimensional relative proper motion. We obtain  $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$   and $\delta \mu_{HST}=0.20\pm 0.10\masyr (47 \pm 24\kms )$.

% Both indicate a non-zero peculiar 2D motion that would make it a slow runaway star, but in both cases the meaurements are also consistent with zero within 2 sigma.

% Figure~\ref{fig:mu_dist} shows the distribution of relative proper motions for other VFTS O, B and WR stars subselecting either those with relative proper motion errors smaller than $0.1\masyr$ (yielding 317 stars) or a more restrictive cut including only those with proper motion errors smaller than $0.05\masyr$ (yielding 36) stars.   We see that the derived Gaia proper motion is very typical for stars in the field.

% \SdM{XXX We need to rethink }

%Therefore, two completely independent measures of the proper motion of VFTS682 relative to the surrounding field, one from HST and one from \emph{Gaia} DR2, yield values of the peculiar three-dimensional velocity of VFTS682 % (\Eqref{eq:speed_around_HST} and
% \Eqref{eq:speed_around_Gaia}, respectively)
%which would make it the most massive runaway star known to date. However, the large errors on both the proper motion measures require confirmation with future astrometric data.

%Subtracting the mean proper motion components given by \citet{lennon:18} from the proper motion of VFTS682 (see \Tabref{tab:vfts682}), we obtain the components of proper motion of the star relative to the surrounding region $\delta\mu_\mathrm{RA}^{Gaia} = 0.10 \pm 0.07\masyr$ and $\delta\mu_\mathrm{DEC}^{Gaia} = 0.08 \pm 0.09\masyr$. These components result in a two-dimensional relative proper motion of $\delta \mu^{Gaia}=0.13\pm 0.09\,\mathrm{mas\ yr^{-1}}$.

% \begin{equation}
%   \label{eq:pm_gaia_around}
%   \delta \mu^{Gaia} = \sqrt{\left(\delta\mu_\mathrm{RA}^{Gaia}\right)^2+\left(\delta\mu_\mathrm{DEC}^{Gaia}\right)^2}
%   = 0.13\pm 0.09\,\mathrm{mas\
%   yr^{-1}} \ \ .
% \end{equation}

% Figure \Figref{fig:pm_polar} shows the distribution in relative proper motion
% of the stars used in \cite{lennon:18} to define the frame of reference
% we also adopt here, cf.~the right panel in their Fig.~2. The
% proper motions are decomposed in the radial and tangential direction
% from R136, to highlight the likelihood of it being the origin of fast
% moving stars. The plus sign marks VFTS682, which sits at the edge of
% the proper motion distribution, but is not a clear outlier. Most
% relevant is the direction of the proper motion of VFTS682, which we
% discuss in \Secref{sec:r136_origin}.

%The proper motion components can be converted into the components of the relative transverse velocity $\delta v_\mathrm{RA}^{Gaia}=24\pm19\,\kms$, $\delta v_\mathrm{DEC}^{Gaia}=20\pm23\,\kms$, assuming a distance of 50\,kpc. These can be combined obtaining a projected two-dimensional velocity of $31\pm21\kms$.

%The radial velocity from \cite{bestenlehner:11} then gives the third component along the line of sight,  which added in quadrature to the transverse components results in a three-dimensional velocity of $44 \pm 21\kms$.
% \begin{equation}
%   \label{eq:speed_around_Gaia}
%   v_\mathrm{pec}^{Gaia} = \sqrt{\left(\delta v_\mathrm{RA}^{Gaia}\right)^2
%     +\left(\delta v_\mathrm{DEC}^{Gaia}\right)^2+\left(\delta
%       v_\mathrm{rad}^{VFTS}\right)^2} = 44 \pm 21
%   \kms \ .
% \end{equation}

% \subsection{Does it come from the R136 cluster?}
% \label{sec:r136_origin}

% Assuming that the best estimate of the peculiar velocity of VFTS682
% are reliable, we now address the question of its likely origin.

% Figure~\ref{fig:mu_dist} shows the distribution in proper motion relative to R136
% of OB-type and Wolf-Rayet stars included both in the VFTS survey and
% \emph{Gaia DR2} with reliable astrometric solutions. The bottom panel
% shows the distribution in angles $\theta$ between the relative proper motion
% direction and the direction from the core of R136 to the star. We
% emphasize that our subset of stars with reliable proper motion
% measurements is biased towards the fast moving objects, which results in a
% distribution of angles mildly peaked at small angles.

% VFTS682 is not an outlier in relative proper motion%  compared
% % to the two O-type runaways (re-)identified by \cite{lennon:18} (see
% % also below)
% . However, the star was expected to be a ``slow runaway'' by \citet{bestenlehner:11} in
% the dynamical ejection scenario. The green shade in
% \Figref{fig:mu_dist} shows the range of relative proper motions
% required to reach the present day location within
% the uncertainties on the apparent age of the star, assuming it was
% ejected from R136 very early in its life.

% Because of the large
% error bars, the angle between the relative proper motion is not very
% constraining, but it is suggestive that the best value is close to zero.
% Therefore, the relative proper motion from \emph{Gaia} DR2 are consistent with the hypothesis
% of dynamical ejection.

% \todo{discuss and find a place for Fig.2}
% The red arrow in \Figref{fig:main} shows the direction of relative proper motion of
% VFTS682 from \emph{Gaia} DR2, and the
% yellow arrows illustrate the uncertainty. % show the possible
% % range of directions within the uncertainties in the measured relative proper
% % motion.
% These arrows cross at the present-day location of VFTS682 and
% are prolonged in the direction opposite to the motion to illustrate
% the possible range of origins.
% The yellow arrow shows the direction of
% the relative proper motion from HST (see \Eqref{eq:pm_around_HST}),
% but we omit to show the possible range of directions from HST data, since it is
% insufficiently constrained to be useful.

% Although the large uncertainties on the relative proper motion
% components results in a wide range of possible directions,we argue that the most likely origin of the star is R136.

 % , which
% corroborates the idea that the star is the result of a dynamical
% ejection very soon in the cluster evolution.

% the thing below is not relevant if it is a merger, better cut out
% for now, since we are too long anyway

%Note that the
% present-day surface helium abundance
% \citep[$Y\simeq0.5$,][]{bestenlehner:11, rubio-diez:17} puts a lower
% limit on the age of the star of $\sim$0.9\,Myr, corresponding to the time needed to
% synthesize this amount of helium in the models from \cite{kohler:15}.

@@

\citet{lennon:18} carried out a study similar to ours on the fast
moving O-type stars
in the region, and found two very massive runaway stars in the 30 Doradus region. One of them (VFTS 16)
was previously known as a runaway star from its line of sight velocity
\citep[][]{evans:10}. \citet{lennon:18} also concluded that VFTS16 is 
the result of a dynamical ejection from the R136 cluster, while the
origin of the other star (VFTS 72) is less clear given its direction
of motion (cf.~\Figref{fig:dist}). The value of $\tau_\mathrm{kin}\simeq0.9$\,Myr we find for
VFTS682 is smaller than the
corresponding value for VFTS16: \cite{lennon:18} inferred a kinematic
age of $\sim$1.5\,Myr, possibly in tension with the apparent age of that star. This means that the more
massive VFTS682 was ejected later than VFTS16 from the same cluster.

>>>>

% Aiming to shine light on this debate,
In this context, 
we investigate the kinematics of
VFTS682 as obtained by \emph{Gaia} and multi-epoch \emph{Hubble Space
  Telescope} photometry. We derive a projected velocity relative to
the cluster of $\sim$$30\pm20\kms$ at one $\sigma$ level.  The direction and magnitude of
the proper motion are consistent with VFTS682 being a slow runaway
ejected from the central cluster, given its distance and the inferred age.  However, the error bars are substantial and the proper motion measurements alone cannot rule out the counter hypothesis with high confidence.

@@

The WNh5 star VFTS682, located at right ascension (RA)
05$^\mathrm{h}$38$^\mathrm{m}$55.510$^\mathrm{s}$  and declination
(DEC) \mbox{-69$^\mathrm{o}$04'26.72''} J2000, was observed as part of the multi-epoch, spectroscopic VFTS campaign covering $\lambda$4000--7000 \citep[][]{evans:11}. 
\citet{bestenlehner:11}  analyzed the spectra to infer the stellar
parameters and measure an extinction of $A_V=4.45\pm0.12$, implying a
luminosity of $\log_{10}(L/L_\odot) =  6.5\pm0.2$, making this one of
the brightest stars in the region. The absence of radial velocity
variations suggests that the star is unlikely to have close binary
companions \citet{bestenlehner:11}, unless the orbital inclination is
very high. Bayesian fits of the stellar
parameters against evolutionary tracks \citep{brott:11, kohler:15}
using the BONNSAI code \citep{schneider:17, schneider:18} provide
estimates for the age, present mass and initial mass, % indicating that
% this star among the most massive stars known,
% see
\Tabref{tab:star_param}. % for an overview of the parameters.

@@

\cite{townsley:06} did not find
an X-ray point source at the position of VFTS682 in their
\emph{Chandra} survey of the region, which suggests the absence of
colliding winds that would be expected in the presence of companions
even for extreme mass ratios, given the extremely large mass of
VFTS682. This star is also relatively isolated in (near) infrared. There nearest bright (near-)infrared sources detected by \emph{Spitzer} \citep{meixner:06} and resolved in the \emph{VISTA} Magellanic Clouds Survey \citep{cioni:11} are located at a distance of about 10 arcsecond, i.e. about 2.4 pc. \cite{walborn:13} speculate that these nearby young stars may represent a case of starformation triggered by the the wind of VFTS682, but we chance alignment cannot be excluded.

@@

We adopt as distance to the LMC the value of $50$\,kpc. The error on
the distance determination is small ($\lesssim2\%$,
\citealt{pietrzynski:13}) and any possible offset in the radial direction between R136 or VFTS682 and the distance we adopted for the LMC is probably much smaller (likely $\ll 1\%$, e.g., \citealt{Luks+1992}). These uncertainties are negligible compared to the errors in the proper motion discussed below.

@@

\vspace*{10pt}% to move new sec. to newpage
\begin{figure*}[tbp]
  \centering
  \includegraphics[width=0.35\textwidth]{figures/dist_mu_region.pdf}
  \includegraphics[width=0.35\textwidth]{figures/angle}
  \caption{\emph{Left panel}: distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136. VFTS682 is not an outlier, but
    its relative proper motion matches the expected value if it were indeed
    ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. The top axis shows the conversion to physical units
    assuming a distance of 50\,kpc. \emph{Right panel}:  distribution of
    angles between the relative proper motion direction and the radial
    direction from the center of the cluster to the star. The error bars from \emph{Gaia} for VFTS682 are large, but
    the best value is in agreement with the hypothesis of dynamical
    ejection. The peak at $\theta\simeq\pi/2$ is due to stars
    belonging to NGC2060. In both
    panels, the dark blue histograms contain 317 
    stars with error smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
    lighter blue histograms contain 36 stars with errors smaller than $0.05\,\mathrm{mas \
      yr^{-1}}$. }
  \label{fig:dist}
\end{figure*}

@@

@@

Assuming VFTS682 indeed originates from R136, we can calculate its kinematic
age simply as:
\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{\delta\mu_{Gaia}} \simeq
  \frac{119.4\,\mathrm{arcsec}}{0.13\masyr} \simeq 0.9\pm\,0.6\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel = 119.4\,\mathrm{arcsec}$ is the angular distance from VFTS682 to
the core of the cluster \citep[corresponding to $\sim$29\,pc at a
distance of 
50\,kpc,][]{bestenlehner:11}.
% As in the rest of this study, we neglect for
% simplicity the error on the distance estimates, because it is negligible compared to other uncertainties.
The kinematic age $\tau_\mathrm{kin}$ is consistent with a very early
ejection from the cluster, given the apparent age of the star.
% of $1.0\pm 0.2$\,Myr 
% \citep{schneider:18}.

@@

@@

In \citet{lennon:18} we carried out a similar study on the fast
moving O-type stars in the region, and found two very massive runaway stars in the 30 Doradus region. One of them (VFTS 16)
was previously known as a runaway star from its line of sight velocity
\citep[][]{evans:10}. \citet{lennon:18} also concluded that VFTS16 is 
the result of a dynamical ejection from the R136 cluster, while the
origin of the other star (VFTS72) is less clear given its direction
of motion (cf.~\Figref{fig:dist}). The value of $\tau_\mathrm{kin}\simeq0.9$\,Myr we find for
VFTS682 is smaller than the
corresponding value for VFTS16 of $\sim$1.5\,Myr, possibly in tension with the apparent age of that star. This means that the more
massive VFTS682 was ejected later than VFTS16 from the same cluster.
